Identifying and Correcting Programming Language Behavior Misconceptions

### 2023-11-14 proposal.tex
grammar check

In the rest of this document, I first define SMoL (\Cref{s:smol}),
then present a literature review of misconceptions
(\Cref{s:background-misconceptions}) and tutoring systems
(\Cref{s:background-tutoring-systems}), and then outline my research
plans for the RQs (\Cref{s:plan-rq1,s:plan-rq2}). After that, I describe preliminary results (\Cref{s:preliminary-results} and propose my next steps (\Cref{s:future-work}).

@@

The Tutor mostly consists of \emph{interpreting tasks}:
multiple-choice questions (MCQs) that asks students to interpret
program. The options include (among other options
that I will explain shortly) the correct answer and wrong answers that
correspond to known misconceptions. Because students might conceive
wrong answers that I do not anticipate, the Tutor has been providing
an ``Other'' option that, once chosen, allows students to enter
arbitrary answers.

@@

To know students well, I would like to have as many tasks per
misconception as possible. But in practice, the number of tasks is
subject to the following limitations:
\begin{enumerate}
  \item I don't want the length of any tutorial section to exceed 30
  minutes.
  \item Programs shouldn't be too long. Some misconceptions are really
  similar and as a consequence have few short programs to tell them
  apart.
\end{enumerate}

@@

The output from Quizius is (a) a collection of programs; (b) for
each program, a collection of predicted answers; and (c) for each
answer, a rationale. Clustering the answers is easy (after
ignoring some small syntactic differences). Thus, for
each cluster, I obtain a set of rationales.

After running Quizius in the course (\Cref{s:populations}),
I took over as experts. Determining which is the right
answer is easy, whereas expert knowledge is useful is in
\emph{clustering the rationales}. If all the rationales for a wrong
answer are fairly similar, this is strong evidence that there is a
common misconception that generates it. If, however, there are
multiple rationale clusters, that means the program
is not discriminative enough to distinguish the misconceptions, and it
needs to be further refined to tell them apart. Interestingly, even
the correct answer needs to be analyzed, because sometimes correct
answers do have incorrect rationales (again, suggesting the program
needs refinement to discriminate correct conceptions
from misconceptions).

@@

  \item I deferred the local binding forms to the end, since I felt
        they were less essential. I therefore rewrote programs that use
        it to not depend on it.

@@

Several groups of authors have argued for the benefits of visualizing
program executions
\citep{dicksonExperiencesImplementingUtilizing2022,
guoOnlinePythonTutor2013, karnalimUsePythonTutor2017,
napsExploringRoleVisualization2002}. Visualization seems to
increase engagement and help students understand how programs work.
These work often mention \emph{notional machines}
\citep{duboulayBlackBoxGlass1981}, which refers to the underlying
mental models rather than the visual representations.

@@

This section describe the current status of the Tutor.

The Tutor includes five tutorials, shown on the left in
\Cref{t:tutor-goal-sentences}. The larger topics are further broken
down into 2--3 modules. The goal was for students to spend at most
20--30 minutes per module. my data show that in practice, students
spent about 9.8 (median) minutes.

@@

In addition to asking students questions, the Tutor along the way
introduces terminology and states the true conceptions. These are the
teaching goals of the Tutor. I therefore refer to these as
\newterm{goal sentences}. \Cref{t:tutor-goal-sentences} lists
the (abbreviated) goal sentences for each tutorial. I plan to keep
goal sentences in the Tutor.

@@

Overall, the data suggest that the Tutor \emph{most likely did not do
harm} and perhaps even \emph{may have done some good}. Revisions on
the Tutor (\Cref{s:future-work}), in light of misinterpreters, is
necessarily to give a more accurate evaluation. For example I found
far fewer problems than I would have liked. Concretely, only 40 of the
71 eligible problems (after removing the non-SMoL modules) were useful
in the above analysis.

@@

% To evaluate,
I computed a Spearman's rank correlation $\rho$, ranking questions by
what percentage of students got the question right. Between the
original university and University 2, we obtain a \pValue~=~2.013e-07.
Between the original university and the online population, I obtain a
\pValue~<~2.2e-16. These show that the other two populations performed
similarly to the original one. While further validation on other
populations remains essential, these suggest that the questions are
\emph{finding misconceptions that may be universal}.

Of course, other threats to generalizability remains: both University
1 and University 2 use the same textbook; SMoL's Lispy syntax might
have bias in probing errors. I plan to address these issues in the
future (\Cref{s:future-work}).

@@

The processes described in
\Cref{s:find-misconceptions-with-quizius,s:find-misconceptions-with-tutor}
have been conducted with SMoL's Lispy syntax (\Cref{s:smol}). This can
potentially introduce biases to the found misconceptions. For example,
I found that some students think the following program
produces \lstinline|'(4 5)| rather than \lstinline|'(1 2)|.
\begin{lstlisting}
  (filter (lambda (n) (> 3 n)) '(1 2 3 4 5))
\end{lstlisting}
This checks whether $\verb|3| > \verb|n|$, but those students
presumably vocalized it as ``greater than \lstinline|3|,
\lstinline|n|?''. Although I have tried to remove programs like this
from the Tutor, there is still a risk of having biases in student
errors. This semester (Fall 2023) I am already deploying a newer SMoL
Tutor at Brown. This new Tutor is able to display programs in
JavaScript and Python, in addition to the Lispy syntax. I plan to
compare in Spring 2024 the new Tutor data with the current results,
i.e., a comparison between uni-syntax Tutor and multi-syntax Tutor.

>>>>

In the rest of this document, I first define SMoL (\Cref{s:smol}),
then present a literature review of misconceptions
(\Cref{s:background-misconceptions}) and tutoring systems
(\Cref{s:background-tutoring-systems}), and then outline my research
plans for the RQs (\Cref{s:plan-rq1,s:plan-rq2}). After that, I describe the preliminary results (\Cref{s:preliminary-results} and propose my next steps (\Cref{s:future-work}).

@@

The Tutor mostly consists of \emph{interpreting tasks}:
multiple-choice questions (MCQs) that ask students to interpret
program. The options include (among other options
that I will explain shortly) the correct answer and wrong answers that
correspond to known misconceptions. Because students might conceive
wrong answers that I do not anticipate, the Tutor has been providing
an ``Other'' option that, once chosen, allows students to enter
arbitrary answers.

@@

To know students well, I would like to have as many tasks per
misconception as possible. But in practice, the number of tasks is
subject to the following limitations:
\begin{enumerate}
  \item I don't want the length of any tutorial section to exceed 30
  minutes.
  \item Programs shouldn't be too long. Some misconceptions are really
  similar and, as a consequence, have few short programs to tell them
  apart.
\end{enumerate}

@@

The output from Quizius is (a) a collection of programs; (b) for
each program, a collection of predicted answers; and (c) for each
answer, a rationale. Clustering the answers is easy (after
ignoring some small syntactic differences). Thus, for
each cluster, I obtained a set of rationales.

After running Quizius in the course (\Cref{s:populations}),
I took over as an expert. Determining which is the right
answer is easy, whereas expert knowledge is useful in
\emph{clustering the rationales}. If all the rationales for a wrong
answer are fairly similar, this is strong evidence that there is a
common misconception that generates it. If, however, there are
multiple rationale clusters, that means the program
is not discriminative enough to distinguish the misconceptions, and it
needs to be further refined to tell them apart. Interestingly, even
the correct answer needs to be analyzed, because sometimes correct
answers do have incorrect rationales (again, suggesting the program
needs refinement to discriminate correct conceptions
from misconceptions).

@@

  \item I deferred the local binding forms to the end, since I felt
        they were less important. I therefore rewrote programs that use
        it to not depend on it.

@@

Several groups of authors have argued for the benefits of visualizing
program executions
\citep{dicksonExperiencesImplementingUtilizing2022,
guoOnlinePythonTutor2013, karnalimUsePythonTutor2017,
napsExploringRoleVisualization2002}. Visualization seems to
increase engagement and help students understand how programs work.
These works often mention \emph{notional machines}
\citep{duboulayBlackBoxGlass1981}, which refers to the underlying
mental models rather than the visual representations.

@@

This section describes the current status of the Tutor.

The Tutor includes five tutorials, shown on the left in
\Cref{t:tutor-goal-sentences}. The larger topics are further broken
down into 2--3 modules. The goal was for students to spend at most
20--30 minutes per module. My data show that in practice, students
spent about 9.8 minutes (median).

@@

In addition to asking students questions, the Tutor, along the way,
introduces terminology and states the true conceptions. These are the
teaching goals of the Tutor. I, therefore, refer to these as
\newterm{goal sentences}. \Cref{t:tutor-goal-sentences} lists
the (abbreviated) goal sentences for each tutorial. I plan to keep
goal sentences in the Tutor.

@@

Overall, the data suggest that the Tutor \emph{most likely did not do
harm} and perhaps even \emph{may have done some good}. Revisions on
the Tutor (\Cref{s:future-work}), in light of misinterpreters, is
necessarily to give a more accurate evaluation. For example, I found
far fewer problems than I would have liked. Concretely, only 40 of the
71 eligible problems (after removing the non-SMoL modules) were useful
in the above analysis.

@@

% To evaluate,
I computed a Spearman's rank correlation $\rho$, ranking questions by
what percentage of students got the question right. Between the
original university and University 2, I obtained a \pValue~=~2.013e-07.
Between the original university and the online population, I obtained a
\pValue~<~2.2e-16. These show that the other two populations performed
similarly to the original one. While further validation on other
populations remains essential, these suggest that the questions are
\emph{finding misconceptions that may be universal}.

Of course, other threats to generalizability remain: both University
1 and University 2 use the same textbook; SMoL's Lispy syntax might
have biases in probing errors. I plan to address these issues in the
future (\Cref{s:future-work}).

@@

The processes described in
\Cref{s:find-misconceptions-with-quizius,s:find-misconceptions-with-tutor}
have been conducted with SMoL's Lispy syntax (\Cref{s:smol}). This can
potentially introduce biases to the found misconceptions. For example,
I found that some students think the following program
produces \lstinline|'(4 5)| rather than \lstinline|'(1 2)|.
\begin{lstlisting}
  (filter (lambda (n) (> 3 n)) '(1 2 3 4 5))
\end{lstlisting}
This checks whether $\verb|3| > \verb|n|$, but those students
presumably vocalized it as ``greater than \lstinline|3|,
\lstinline|n|?''. Although I have tried to remove programs like this
from the Tutor, there is still a risk of having biases in student
errors. This semester (Fall 2023), I am already deploying a newer SMoL
Tutor at Brown. This new Tutor is able to display programs in
JavaScript and Python, in addition to the Lispy syntax. I plan to
compare in Spring 2024 the new Tutor data with the current results,
i.e., a comparison between uni-syntax Tutor and multi-syntax Tutor.
